$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers

### 2020-10-18 src/s_gbm_vs_rf.Rmd
first draft of gbm vs rf sections

r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(1000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))

r_star_opts <- tibble(d=m_parameters$ds, nu2=m_parameters$nu2, optimal=r_star_opts1)
temp_df <- 
  big_df %>%
  left_join(r_star_opts) %>% 
  select(-contains("time")) %>% 
  select(-one_of("nu1")) %>% 
  melt(id.vars=c("iter", "d", "nu2")) %>% 
  group_by(variable, d, nu2) %>% 
  summarise(lower=quantile(value, 0.25),
            middle=quantile(value, 0.5),
            upper=quantile(value, 0.75)) %>% 
  ungroup() %>% 
  mutate(variable=as.character(variable)) %>% 
  mutate(variable=if_else(variable=="rstar_rf", "rf", if_else(variable=="rstar_gbm", "gbm", "optimal")))
saveRDS(temp_df, "../data/gbm_vs_rf_studentts_results_optimal.rds")
temp_df <- readRDS("../data/gbm_vs_rf_studentts_results_optimal.rds") %>% 
  mutate(variable=as.character(variable)) %>% 
  mutate(variable=if_else(variable=="gbm", "GBM",
                          if_else(variable=="rf", "RF", "Optimal"))) %>% 
  mutate(variable=fct_relevel(as.factor(variable), "GBM", "RF", "Optimal"))
g2 <- ggplot(data=temp_df,
       aes(x=as.factor(nu2), y=middle, colour=as.factor(variable),
           shape=as.factor(variable))) +
  geom_pointrange(aes(ymin=lower, ymax=upper),
                  position = position_dodge2(width=0.3)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_color_grey("Method", start=0.8, end=0) +
  scale_shape("Method") +
  xlab("# degrees of freedom") +
  ylab("R*") +
  theme(text = element_text(size=16, colour="black"),
        axis.text = element_text(colour="black")) +
  facet_wrap(~d)
ggsave("../output/gbm_vs_rf_studentt.pdf", g2, width = 12, height = 8)
```

>>>>

r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(10000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))

r_star_opts <- tibble(d=m_parameters$ds, nu2=m_parameters$nu2, optimal=r_star_opts1)
big_df <- readRDS("../data/gbm_vs_rf_studentts_results.rds")
temp_df <- 
  big_df %>%
  left_join(r_star_opts) %>% 
  select(-contains("time")) %>% 
  select(-one_of("nu1")) %>% 
  melt(id.vars=c("iter", "d", "nu2")) %>% 
  group_by(variable, d, nu2) %>% 
  summarise(lower=quantile(value, 0.25),
            middle=quantile(value, 0.5),
            upper=quantile(value, 0.75)) %>% 
  ungroup() %>% 
  mutate(variable=as.character(variable)) %>% 
  mutate(variable=if_else(variable=="rstar_rf", "rf", if_else(variable=="rstar_gbm", "gbm", "optimal")))
saveRDS(temp_df, "../data/gbm_vs_rf_studentts_results_optimal.rds")
temp_df <- readRDS("../data/gbm_vs_rf_studentts_results_optimal.rds") %>% 
  mutate(variable=as.character(variable)) %>% 
  mutate(variable=if_else(variable=="gbm", "GBM",
                          if_else(variable=="rf", "RF", "Optimal"))) %>% 
  mutate(variable=fct_relevel(as.factor(variable), "GBM", "RF", "Optimal"))
g2 <- ggplot(data=temp_df,
       aes(x=as.factor(nu2), y=middle, colour=as.factor(variable),
           shape=as.factor(variable))) +
  geom_pointrange(aes(ymin=lower, ymax=upper),
                  position = position_dodge2(width=0.3)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_color_grey("Method", start=0.8, end=0) +
  scale_shape("Method") +
  xlab("# degrees of freedom") +
  ylab("R*") +
  theme(text = element_text(size=16, colour="black"),
        axis.text = element_text(colour="black")) +
  facet_wrap(~d)
ggsave("../output/gbm_vs_rf_studentt.pdf", g2, width = 12, height = 8)
```

### 2020-10-18 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
first draft of gbm vs rf sections

\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
In this section, we illustrate how $R^*$ can diagnose convergence issues in the joint target distribution.

@@

\section{Acknowledgements}
The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.

>>>>

\color{red}
\subsection{Comparing GBMs and RFs}
In section SX, we perform a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them. In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier. In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}. In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution. In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains. Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}. This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw. Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.

\subsubsection{Joint distribution}\label{sec:joint_distribution}
Here, we consider a multivariate normal target distribution over a range of target dimensions from 1-32. For three `chains', we generate 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix. To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals. To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$. For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers. In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and determine the predictive accuracy using the Bayes optimal classifier.

The results of these experiments are shown in Fig. \ref{fig:gbm_vs_rf_normal}. In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic. When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$. In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence. As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows. $R^*$ as calculated by GBMs and RFs also increases. For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification. For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares). The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians. For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process. The dashed line shows $R^*=1$. Here, $R^*$ was calculated using chains split into two halves.}
	\label{fig:gbm_vs_rf_normal}
\end{figure}

\subsubsection{Fat tails}\label{sec:tail_fatness}
We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution. To do so, we again consider four chains. In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom. In the remaining chain, we generate the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$. We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$. We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness. Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$. To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$. For each combination of dimensions and $\nu$, we perform 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers. Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.

The results of these experiments are shown in Fig. \ref{fig:gbm_vs_rf_studdentt}. Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions. Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$. The point colour and shadings give the method used to calculate $R^*$. In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method. As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases. In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers. In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate. If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.

Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains. The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy. In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness. These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones. This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and section X (hyperparameters)): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time. Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time. RFs are less sensitive to hyperparameter variation (section X), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}. Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution. The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares). The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians. For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process. The dashed line shows $R^*=1$. Here, $R^*$ was calculated using chains split into two halves.}
	\label{fig:gbm_vs_rf_studdentt}
\end{figure}

\color{black}
\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
In this section, we illustrate how $R^*$ can diagnose convergence issues in the joint target distribution.

@@

\section{Acknowledgements}
\textcolor{red}{The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.}
